Cloudbreak: Accurate and Scalable Genomic Structural Variation Detection in the Cloud with MapReduce \\ Supplementary Material

### 2014-01-24 manuscript/bioinfo/document.tex
comments from kemal

\title[Cloudbreak]{{\em Cloudbreak}: Accurate and Scalable Genomic Structural Variation Detection in the Cloud with MapReduce}
\author[Whelan \textit{et~al}]{Christopher W. Whelan\,$^{1}$, Jeffrey Tyner\,$^{2}$, Alberto L'Abbate\,$^{3}$, Clelia Tiziana Storlazzi,$^{3}$, Lucia Carbone\,$^{4}$, and Kemal S\"onmez\,$^{1,5,*}$\footnote{to whom correspondence should be addressed}}
\address{$^{1}$Institute on Development and Disability and Center for Spoken Language Understanding, Oregon Health \& Science University, Portland, OR, USA \\
$^{2}$ Program in Molecular and Cellular Biosciences, Oregon Health \& Science University, Portland, OR, USA \\
$^{3}$ Department of Biology, University of Bari ``Aldo Moro'', Via G. Amendola 165/A, 70126, Bari, Italy \\
$^{4}$ Behavioral Neuroscience Department, Oregon Health \& Science University, Portland, OR, USA \\
$^{5}$ Department of Medical Informatics \& Clinical Epidemiology, Oregon Health \& Science University, Portland, OR, USA 
}

@@

\section{Availability:}
Source code, binary releases, example scripts, and a full user manual are available at \url{http://github.com/cwhelan/cloudbreak}

@@

\begin{figure}
\centering
\includegraphics[width=.5\textwidth]{../../figures/workflow_with_whirr.png}
\caption{An overview of the Cloudbreak workflow. Reads are first uploaded to a Hadoop cluster from local storage. Cloudbreak then executes three MapReduce jobs to process the data: 1) Mapping with sensitive settings. 2) Computation of features across the genome. 3) Calling structural variations based on the features computed in the previous step. Finally, SV predictions can be downloaded from the Hadoop cluster and examined. Cloudbreak can also use the Apache Whirr library to automatically provision Hadoop clusters on and deploy data to cloud providers such as Amazon Elastic Compute Cloud.}
\label{cloudbreak_workflow}
\end{figure}

@@

We have implemented an RP-based SV detection package, Cloudbreak, in the framework described above. Cloudbreak detects deletions of size 40bp to 25kb, and small insertions. We accomplish this by providing the following implementions of the user-defined functions described above:

@@

\begin{figure*}[!tpb]
\centering
\includegraphics[width=1\textwidth]{../../figures/chr2_rocs_runtimes_journal.pdf}
\caption{Accuracy and runtime performance on a simulated data set. (a) Receiver Operating Characteristic (ROC) curves showing the specificity and sensitivity of each tool to deletions larger than 40bp on a simulated set of reads giving diploid coverage of 30X on human chromosome 2. Deletions from the Venter genome were randomly added to one or both haplotypes. Each point on a curve represents a different threshold on the confidence of predictions made by that tool. Thresholds vary by: Cloudbreak - likelihood ratio; BreakDancer, DELLY, GASVPro - number of supporting read pairs; Pindel - simple score. (b) ROC curves for insertion predictions. (c) Runtimes for each tool, not including alignment time, parallelized when possible (see Supplementary Material).}
\label{chr2CombinedRoc}
\end{figure*}

Figure \ref{chr2CombinedRoc} shows Receiver Operating Characteristic (ROC) curves of the performance of each algorithm at identifying variants larger than 40bp on the simulated data set, as well as the runtimes of the approaches tested, excluding alignment. See Supplementary Material for a description of how we identified correct predictions. All approaches show high specificity at stringent thresholds in this simulation. Cloudbreak provides the greatest specificity for deletions at higher levels of sensitivity, followed by DELLY. For insertions, Cloudbreak's performance is similar to Pindel, although Cloudbreak can achieve greater sensitivity at the same false positive rate by decreasing its threshold. Cloudbreak's runtime is half that of BreakDancer, the next fastest tool, processing the simulated data in under six minutes. (Of course, Cloudbreak uses many more CPUs as distributed algorithm. See Supplementary Material for a discussion of runtimes and parallelization.) The output which we obtained from MoDIL did not have a threshold that could be varied to correlate with the trade-off between precision and recall and therefore it is not included in ROC curves; in addition, MoDIL ran for 52,547 seconds using 250 CPUs in our cluster. Apart from the alignment phase, which is embarrassingly parallel, the feature generation job is the most computationally intensive part of the Cloudbreak workflow. To test its scalability we measured its runtime on Hadoop clusters made up of varying numbers of nodes and observed that linear speedups can be achieved in this portion of the algorithm by adding additional nodes to the cluster until a point of diminishing returns is reached (Supplementary Fig. \ref{Sscalability}).

@@

It should be noted that the methods tested here vary in their breakpoint resolution (Supplementary Fig. \ref{Sbreakpoint_resolution}). SR methods have higher resolution than RP methods. Cloudbreak's resolution suffers because the independent calculation of features at each window from the distribution of overlapping insert sizes does not allow the algorithm to keep track of the actual coordinates of the mapped reads from each pair, which other tools use to set bounds on the true breakpoint locations. We believe, however, that it is worth increasing sensitivity and specificity in SV detection tools even at the cost of breakpoint resolution. \todo{It would also be possible to extend our feature set to include information useful for increasing resolution, such as the location of nearby read mappings or soft-clipped alignments. }In addition, several recent pipelines have validated RP calls \emph{in silico} by local assembly, a process that could potentially be parallelized in a subsequent MapReduce job.

@@

\begin{figure*}
\centering
\includegraphics[width=1\textwidth]{../../figures/NA18507_rocs_runtimes_journal.pdf}
\caption{Accuracy and performance on the 37X NA18507 sample. (a) ROC curve for deletion prediction performance, tested against the combined gold standard sets of deletions taken from~\cite{Kidd:2008p926},~\cite{Mills:2011fi}, and~\cite{GenomesProjectConsortium:2012co}. (b) ROC curve for insertion prediction performance. (c) Runtimes for each tool, not including alignment time, parallelized when possible (see Supplementary Material). }
\label{NA18507CombinedRoc}
\end{figure*}

@@

We also tested Cloudbreak on a sequencing data set obtained from a patient with acute myeloid leukemia (AML). This data set consisted of 76bp paired end reads with a mean insert size of 285bp and standard deviation of 50bp, yielding sequence coverage of 5X and physical coverage of 8X. Using a pipeline consisting of Novoalign, BreakDancer, and a set of custom scripts for filtering and annotating candidate SVs, we had previously identified a set of variants present in this sample and validated several using PCR, including 8 deletions. Cloudbreak was able to identify all 8 of the validated deletions, showing that it is still sensitive to variants even when using lower coverage data sets with a greater variance of insert sizes. The variants identified include deletions in the gene CTDSPL/RBPS3, an AML tumor suppressor~\citep{Zheng:2012kk}, and NBEAL1, a gene up-regulated in some cancers~\citep{Chen:2004jo}. We are currently investigating these deletions to determine their functional impact on this patient.

@@

Over the next few years, due to advances in sequencing technology, genomics data are expected to increase in volume by several orders of magnitude, expanding into the realm referred to as ``big data''. In addition, the usage of existing information will increase drastically as research in genomics grows and translational applications are developed, and the data sets are reprocessed and integrated into new pipelines. In order to capitalize on this emerging wealth of genome data, novel computational solutions that are capable of scaling with the increasing number and size of these data sets will have to be developed.

>>>>

\title[Cloudbreak]{Cloudbreak: Accurate and Scalable Genomic Structural Variation Detection in the Cloud with MapReduce}
\author[Whelan \textit{et~al}]{Christopher W. Whelan\,$^{1}$, Jeffrey Tyner\,$^{2}$, Alberto L'Abbate\,$^{3}$, Clelia Tiziana Storlazzi,$^{3}$, Lucia Carbone\,$^{4}$, and Kemal S\"onmez\,$^{1,5,*}$\footnote{to whom correspondence should be addressed}}
\address{$^{1}$Institute on Development and Disability and Center for Spoken Language Understanding, Oregon Health \& Science University (OHSU), Portland, OR, USA \\
$^{2}$ Program in Molecular and Cellular Biosciences, OHSU\todo{Oregon Health \& Science University}, Portland, OR, USA \\
$^{3}$ Department of Biology, University of Bari ``Aldo Moro'', Via G. Amendola 165/A, 70126, Bari, Italy \\
$^{4}$ Behavioral Neuroscience Department, OHSU\todo{Oregon Health \& Science University}, Portland, OR, USA \\
$^{5}$ Department of Medical Informatics \& Clinical Epidemiology, OHSU\todo{Oregon Health \& Science University}, Portland, OR, USA 
}

@@

\section{Availability:}
Source code, binary releases, and user manual are available at \url{http://github.com/cwhelan/cloudbreak}

\section{Supplementary information:} Supplementary material is available at \emph{Bioinformatics} online.

@@

\begin{figure}
\centering
\includegraphics[width=.5\textwidth]{workflow_with_whirr.png}
\caption{An overview of the Cloudbreak workflow. Reads are first uploaded to a Hadoop cluster from local storage. Cloudbreak then executes three MapReduce jobs to process the data: 1) Mapping with sensitive settings. 2) Computation of features across the genome. 3) Calling structural variations based on the features computed in the previous step. Finally, SV predictions can be downloaded from the Hadoop cluster and examined. Cloudbreak can also use the Apache Whirr library to automatically provision Hadoop clusters on and deploy data to cloud providers such as Amazon Elastic Compute Cloud.}
\label{cloudbreak_workflow}
\end{figure}

@@

We have implemented an RP-based SV detection package, Cloudbreak, in the framework described above. Cloudbreak detects deletions of size 40bp to 25kb, and small insertions. We accomplish this by providing the following implementations of the user-defined functions described above:

@@

\begin{figure*}[!tpb]
\centering
\includegraphics[width=1\textwidth]{chr2_rocs_runtimes_journal.pdf}
\caption{Accuracy and runtime performance on a simulated data set. (a) Receiver Operating Characteristic (ROC) curves showing the specificity and sensitivity of each tool to deletions larger than 40bp on a simulated set of reads giving diploid coverage of 30X on human chromosome 2. Deletions from the Venter genome were randomly added to one or both haplotypes. Each point on a curve represents a different threshold on the confidence of predictions made by that tool. Thresholds vary by: Cloudbreak - likelihood ratio; BreakDancer, DELLY, GASVPro - number of supporting read pairs; Pindel - simple score. (b) ROC curves for insertion predictions. (c) Runtimes for each tool, not including alignment time, parallelized when possible (see Supplementary Material).}
\label{chr2CombinedRoc}
\end{figure*}

Figure \ref{chr2CombinedRoc} shows Receiver Operating Characteristic (ROC) curves of the performance of each algorithm at identifying variants larger than 40bp on the simulated data set, as well as the runtimes of the approaches tested, excluding alignment. See Supplementary Material for a description of how we identified correct predictions. All approaches show high specificity to deletions at stringent thresholds in this simulation. When specificity is relaxed slightly, Cloudbreak provides the highest levels of sensitivity, followed by DELLY. For insertions, Cloudbreak's performance is similar to Pindel, although Cloudbreak can achieve greater sensitivity at the same false positive rate by decreasing its threshold. Cloudbreak's runtime is half that of BreakDancer, the next fastest tool, processing the simulated data in under six minutes. (Of course, Cloudbreak uses many more CPUs as distributed algorithm. See Supplementary Material for a discussion of runtimes and parallelization.) The output which we obtained from MoDIL did not have a threshold that could be varied to correlate with the trade-off between precision and recall and therefore it is not included in ROC curves; in addition, MoDIL ran for 52,547 seconds using 250 CPUs in our cluster. Apart from the alignment phase, which is embarrassingly parallel, the feature generation job is the most computationally intensive part of the Cloudbreak workflow. To test its scalability we measured its runtime on Hadoop clusters made up of varying numbers of nodes and observed that linear speedups can be achieved in this portion of the algorithm by adding additional nodes to the cluster until a point of diminishing returns is reached (Supplementary Fig. \ref{Sscalability}).

@@

We designed Cloudbreak to optimize for sensitivity and specificity for SV events, as demonstrated in the ROC curves, as opposed to attempting to achieve high levels of breakpoint resolution. Supplementary Fig. \ref{Sbreakpoint_resolution} displays the breakpoint resolution of each of the methods tested on the simulated data set. Cloudbreak's resolution is lower than other methods because the independent calculation of features at each window does not allow the algorithm to keep track of the actual coordinates of the mapped reads from each pair, which other tools use to set bounds on the true breakpoint locations. In practice, RP-based predictions are typically validated using techniques such as targeted split read search, local assembly, or wet-lab methods, processes which refine their breakpoint predictions. Therefore, Cloudbreak attempts to maximize the accurate detection of the presence of SV events.

@@

\begin{figure*}
\centering
\includegraphics[width=1\textwidth]{NA18507_rocs_runtimes_journal.pdf}
\caption{Accuracy and performance on the 37X NA18507 sample. (a) ROC curve for deletion prediction performance, tested against the combined gold standard sets of deletions taken from~\cite{Kidd:2008p926},~\cite{Mills:2011fi}, and~\cite{GenomesProjectConsortium:2012co}. (b) ROC curve for insertion prediction performance. (c) Runtimes for each tool, not including alignment time, parallelized when possible (see Supplementary Material). }
\label{NA18507CombinedRoc}
\end{figure*}

@@

We also tested Cloudbreak on a data set obtained from a patient with acute myeloid leukemia (AML). This data consisted of 76bp paired end reads with a mean insert size of 285bp and standard deviation of 50bp, yielding sequence coverage of 5X and physical coverage of 8X. Using a pipeline consisting of Novoalign, BreakDancer, and a set of custom scripts for filtering and annotating candidate SVs, we had previously identified a set of variants present in this sample and validated several using PCR, including 8 deletions. Cloudbreak was able to identify all of the validated deletions, showing that it is still sensitive even when using lower coverage data with a greater variance of insert sizes. The variants identified include deletions in the gene \emph{CTDSPL}/\emph{RBPS3}, an AML tumor suppressor~\citep{Zheng:2012kk}, and NBEAL1, a gene up-regulated in some cancers~\citep{Chen:2004jo}. We are currently investigating these SVs to determine their functional impact on this patient.

@@

Over the next few years, due to advances in sequencing technology, genomics data are expected to increase in volume by several orders of magnitude, expanding into the realm referred to as ``big data''. In addition, the usage of existing information will increase drastically as research in genomics grows and translational applications are developed, and data is reprocessed and integrated into new pipelines. In order to capitalize on this emerging wealth of genome data, novel computational solutions that are capable of scaling with the increasing number and size of these data sets will have to be developed.

### 2014-01-24 manuscript/bioinfo/supplementary.tex
comments from kemal

\title{{\em Cloudbreak}: Accurate and Scalable Genomic Structural Variation Detection in the Cloud with MapReduce \\
Supplementary Material}

@@

\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{../../figures/insert_size_mixtures.pdf}
\caption{Illustration of insert size mixtures at individual genomic locations. A) There is no variant present at the location indicated by the vertical line (left), so the mix of insert sizes (right) follows the expected distribution of the library centered at 200bp, with a small amount of noise coming from low-quality mappings. B) A homozygous deletion of 50bp at the location has shifted the distribution of observed insert sizes. C) A heterozygous deletion at the location causes a mixture of normal and long insert sizes to be detected. D) A heterozygous small insertion shifts a portion of the mixture to have lower insert sizes.}
\label{insert_size_mixes}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{../../figures/cloudbreak_mapred_diagram.pdf}
\caption{An example of the Cloudbreak MapReduce algorithm. A) In the first MapReduce job, mappers scan input reads in FASTQ format and execute an alignment program in either paired-end or single-ended mode to generate read mappings. Reducers gather all alignments for both reads in each pair. B) In the second MapReduce job, mappers first emit information about each read pair (in this case the insert size and quality) under keys indicating the genomic location spanned by that pair. Only one genomic location is diagrammed here for simplicity. Reducers then compute features for each location on the genome by fitting a GMM to the distribution of spanning insert sizes. C) Mappers group all emitted features by their chromosome, and reducers find contiguous blocks of features that indicate the presence of a deletion.}
\label{algorithm_example}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{../../figures/CHR2SIM_ROCS_MULTIPLE_MAPPINGS.pdf}
\caption{Cloudbreak performance on the chromosome 2 simulation using different alignment strategies. ROC curves show the number of true positives and false positives for each operating point for deletions and insertions. The Cloudbreak alignment strategies are: 1) ``Cloudbreak'': Alignments generated with BWA in paired-end mode, reporting the best hit for each pair. 2) ``Cloudbreak-BWA-PE-MM'': Alignments generated with BWA in paired-end mode, reporting up to 25 additional hits for each mapping in SAM format using the \texttt{-n} and \texttt{-N} parameters for \texttt{bwa sampe} and the script \texttt{xa2multi.pl}. 3) ``Cloudbreak-GEM-SE-MM'': Alignments generated by running the GEM aligner in single-ended mode, reporting up to 1000 additional hits per alignment. GEM was executed in parallel using Hadoop tasks which wrap GEM version 1.362 (beta), with parameters \texttt{-e 6 -m 6 -s 2 -q ignore -d 1000 --max-big-indel-length 0},  requesting all hits for a read that are within an edit distance of 6 of the reference, within 2 strata of the best hit, with a maximum of 1000 possible alignments reported for each read. Considering multiple mappings improves Cloudbreak's specificity for insertions but decreases sensitivity to deletions.}
\label{alignment_comparison}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{../../figures/runtimeByNodes.pdf}
\caption{Scalability of the Cloudbreak algorithm. Runtime of the Cloudbreak feature generation job for the simulated Chromosome 2 data is shown on Hadoop clusters consisting of varying numbers of compute nodes. Clusters were created in the Amazon Elastic Compute Cloud.}
\label{scalability}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{../../figures/breakpoint_resolution.pdf}
\caption{Breakpoint resolution for each tool for deletions on the Chromosome 2 simulated data. For each correctly predicted deletion, we calculated the difference in length between the true deletion and the prediction.}
\label{breakpoint_resolution}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=.8\textwidth]{../../figures/runtime_by_windowSize.pdf}
\caption[Cloudbreak runtimes with varying window sizes.]{Runtimes for Cloudbreak on the Chromosome 2 simulated data set using differing choices of window size. Runtimes include the feature computation and variant calling Cloudbreak Hadoop jobs.}
\label{figure_runtime_by_window_size}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=.8\textwidth]{../../figures/breakpoint_resolution_by_windowSize.pdf}
\caption[Cloudbreak breakpoint resolution with varying window sizes.]{Breakpoint resolution for Cloudbreak on the Chromosome 2 simulated data set using differing choices of window size.}
\label{breakpoint_resolution_by_windowSize}
\end{figure}

@@

To handle incorrect and ambiguous mappings, we assume that in general they will not form normally distributed clusters in the same way that correct mappings will, and therefore use an outlier detection technique to filter the observed insert sizes for each location. We sort the observed insert sizes and define as an outlier an observation whose $k$th nearest neighbor, using differences between insert sizes as a measure of distance, is more than $n\sigma$ distant, where $k = 3$ and $n = 5$. In addition, we rank all observations by the estimated probability that the mapping is correct and use an \emph{adaptive quality cutoff} to filter observations: we discard all observations where the estimated probability the mapping is correct is less than the score of the maximum quality observation minus a constant $c$. This allows the use of more uncertain mappings in repetitive regions of the genome while restricting the use of low-quality mappings in unique regions. Defining $\textsc{Mismatches}(a)$ to be the number of mismatches between a read and the reference genome in the alignment $a$, we approximate the probability $p^{k}_c$ of each end alignment being correct by:

@@

And then multiply $p_c(a^{1}_{m,i})$ and $p_c(a^{2}_{m,i})$ to approximate the likelihood that the pair is mapped correctly.

@@

We ran Cloudbreak with a median filter window size of 5 genomic windows.

>>>>

\title{Cloudbreak: Accurate and Scalable Genomic Structural Variation Detection in the Cloud with MapReduce \\
Supplementary Material}

@@

\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{insert_size_mixtures.pdf}
\caption{Illustration of insert size mixtures at individual genomic locations. A) There is no variant present at the location indicated by the vertical line (left), so the mix of insert sizes (right) follows the expected distribution of the library centered at 200bp, with a small amount of noise coming from low-quality mappings. B) A homozygous deletion of 50bp at the location has shifted the distribution of observed insert sizes. C) A heterozygous deletion at the location causes a mixture of normal and long insert sizes to be detected. D) A heterozygous small insertion shifts a portion of the mixture to have lower insert sizes.}
\label{insert_size_mixes}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{cloudbreak_mapred_diagram.pdf}
\caption{An example of the Cloudbreak MapReduce algorithm. A) In the first MapReduce job, mappers scan input reads in FASTQ format and execute an alignment program in either paired-end or single-ended mode to generate read mappings. Reducers gather all alignments for both reads in each pair. B) In the second MapReduce job, mappers first emit information about each read pair (in this case the insert size and quality) under keys indicating the genomic location spanned by that pair. Only one genomic location is diagrammed here for simplicity. Reducers then compute features for each location on the genome by fitting a GMM to the distribution of spanning insert sizes. C) Mappers group all emitted features by their chromosome, and reducers find contiguous blocks of features that indicate the presence of a deletion.}
\label{algorithm_example}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{CHR2SIM_ROCS_MULTIPLE_MAPPINGS.pdf}
\caption{Cloudbreak performance on the chromosome 2 simulation using different alignment strategies. ROC curves show the number of true positives and false positives for each operating point for deletions and insertions. The Cloudbreak alignment strategies are: 1) ``Cloudbreak'': Alignments generated with BWA in paired-end mode, reporting the best hit for each pair. 2) ``Cloudbreak-BWA-PE-MM'': Alignments generated with BWA in paired-end mode, reporting up to 25 additional hits for each mapping in SAM format using the \texttt{-n} and \texttt{-N} parameters for \texttt{bwa sampe} and the script \texttt{xa2multi.pl}. 3) ``Cloudbreak-GEM-SE-MM'': Alignments generated by running the GEM aligner in single-ended mode, reporting up to 1000 additional hits per alignment. GEM was executed in parallel using Hadoop tasks which wrap GEM version 1.362 (beta), with parameters \texttt{-e 6 -m 6 -s 2 -q ignore -d 1000 --max-big-indel-length 0},  requesting all hits for a read that are within an edit distance of 6 of the reference, within 2 strata of the best hit, with a maximum of 1000 possible alignments reported for each read. Considering multiple mappings improves Cloudbreak's specificity for insertions but decreases sensitivity to deletions.}
\label{alignment_comparison}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{runtimeByNodes.pdf}
\caption{Scalability of the Cloudbreak algorithm. Runtime of the Cloudbreak feature generation job for the simulated Chromosome 2 data is shown on Hadoop clusters consisting of varying numbers of compute nodes. Clusters were created in the Amazon Elastic Compute Cloud.}
\label{scalability}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{breakpoint_resolution.pdf}
\caption{Breakpoint resolution for each tool for deletions on the Chromosome 2 simulated data. For each correctly predicted deletion, we calculated the difference in length between the true deletion and the prediction.}
\label{breakpoint_resolution}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=.8\textwidth]{runtime_by_windowSize.pdf}
\caption[Cloudbreak runtimes with varying window sizes.]{Runtimes for Cloudbreak on the Chromosome 2 simulated data set using differing choices of window size. Runtimes include the feature computation and variant calling Cloudbreak Hadoop jobs.}
\label{figure_runtime_by_window_size}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=.8\textwidth]{breakpoint_resolution_by_windowSize.pdf}
\caption[Cloudbreak breakpoint resolution with varying window sizes.]{Breakpoint resolution for Cloudbreak on the Chromosome 2 simulated data set using differing choices of window size.}
\label{breakpoint_resolution_by_windowSize}
\end{figure}

@@

To handle incorrect and ambiguous mappings, we assume that in general they will not form normally distributed clusters in the same way that correct mappings will, and therefore use an outlier detection technique to filter the observed insert sizes for each location. We sort the observed insert sizes and define as an outlier an observation whose $k$th nearest neighbor, using differences between insert sizes as a measure of distance, is more than $n\sigma$ distant, where $k = 3$ and $n = 5$. The parameter $n=5$ represents the choice that only extreme outlier sets are to be eliminated and the parameter $k=3$ represents a trade-off in selecting size of the outlier set between eliminating true modes and allowing small outlier groups to skew the inference, which has been found by observing the distribution of outliers in data.

In addition, when the input data contains multiple possible mappings for each read pair, we calculate a score based on the probability that each mapping is correct relative to the other possible mappings. We then rank all observations at each genomic location, which in most cases include mappings from multiple read pairs, by their scores and use an \emph{adaptive quality cutoff} to filter observations: we discard all observations where the score is less than the maximum score observed at that location minus a constant $c$. This allows the use of more uncertain mappings in repetitive regions of the genome while restricting the use of low-quality mappings in unique regions. We compute mapping scores as follows: defining $\textsc{Mismatches}(a)$ to be the number of mismatches between a read and the reference genome in the alignment $a$, we compute the score $p^{k}_c$ of each end alignment by:

@@

Where $k$ is the position of the read in the pair (i.e. 1,2), $m$ is the identifier of the read pair, and $i$ is the index of a particular possible mapping in the set of all possible mappings for that read. We then multiply $p_c(a^{1}_{m,i})$ and $p_c(a^{2}_{m,i})$ to give a score for that read pair. This score calculation was designed to give higher scores to the set of alignments with the smallest number of mismatches, while dividing smaller amounts of score among possible alignments with more mismatches than the best alignment in the set. When the data only contains one possible mapping for each read pair, as is the case in the experiments reported in the main manuscript, all read pairs receive a score of 1 under this scheme and therefore the adaptive quality cutoff is not invoked.

@@

We ran Cloudbreak with a median filter window size of 5 genomic windows, an adaptive quality cutoff of 6, and a minimum coverage after filtering at each window of 3 spanning read pairs (parameters \texttt{--maxLogMapqDiff 6 --minCleanCoverage 3} on the feature generation step and \texttt{--medianFilterWindow 5} on the variant extraction steps). We chose these parameters based on limited tuning on the simulated data set.
